Combinatorial Proofs and Decomposition Theorems for First-order Logic\vspace{-3ex}

### 2021-01-25 main.tex
multiple changes

\newcommand{\compl}[1]{#1^\complement}
%\newcommand{\sublist}[1]{\set{#1}}
%\newcommand{\subst}[2]{#1\mathord{\scriptstyle\mapsto} #2}
\newcommand{\sublist}[1]{[#1]}
\newcommand{\subst}[2]{#1/#2}
\newcommand{\ssubst}[2]{\sublist{\subst{#1}{#2}}}
\newcommand{\substof}[1]{\sigma_{\!#1}}
\newcommand{\rsubstof}[1]{\rho_{#1}}
\newcommand{\dsubstof}[1]{\delta_{#1}}
\newcommand{\linkingof}[1]{\sim_{#1}}
\newcommand{\linking}{\sim}
\newcommand{\mapof}[1]{\lfloor{#1}\rfloor}
\newcommand{\labelof}[1]{\ell(#1)}
\newcommand{\form}[1]{\mathrm{fm}(#1)}
\newcommand{\formtree}[1]{\mathcal{F}(#1)}

@@

In this paper we advance the idea that combinatorial proofs can provide
a notion of proof identity for first-order logic. \emph{First-order
combinatorial proofs} were introduced by Hughes in~\cite{hughes:fopws}.
For example, a first-order combinatorial proof of Smullyan's 
``drinker paradox'' $\drinkerformula=\veedrinkerformula$ is shown on the
right of Fig.\,\ref{fig:translate},
a homomorphism from a partially coloured graph to a labelled graph.
However, even though Hughes proves soundness and completeness, the 
proof is highly unsatisfactory: (1) the soundness argument is extremely
long, intricate and cumbersome, and (2) the completeness proof does not 
allow a syntactic proof to be read back from a combinatorial proof, i.e.,
completeness is not
\emph{sequentializable} \cite{girard:87} or
\emph{full}~\cite{abramsky:jagadeesan:94}.\dominic{ok to say sequentializable too?}
%
A fundamental problem is that not all combinatorial
proofs can be obtained as translations of sequent calculus proofs.

In this paper we solve these issues by moving to a deep inference
system. More precisely, we introduce a new proof system, called
$\FOKS$, for first-order logic, that (a) reflects every combinatorial
proof, i.e., there is a surjection from $\FOKS$ proofs to
combinatorial proofs, (b) yields far simpler proofs of
soundness and completeness of combinatorial proofs, and (c) admits
new decomposition theorems establishing a precise correspondence
between certain syntactic inference rules and certain combinatorial
notions.
%
The right side of Fig.\,\ref{fig:translate}
illustrates the surjection in (a), and since the syntactic proofs 
of the two systems both translate the same combinatorial proof,
they can be considered \emph{the same}.

In general, a \emph{decomposition theorem} provides normal forms of
proofs, separating subsets of inference rules of a proof system. A
prominent example of a decomposition theorem is Herbrand's
theorem~\cite{herbrand:phd}, which allows a separation between the
propositional part and the quantifier part in a first-order
proof~\cite{gentzen:35:II,brunnler:06:herbrand}. Through the advent of
deep inference, new kinds of proof decompositions became possible,
most notably the separation between the linear part of a proof and the
resource management of a proof. It has been shown by
Stra{\ss}burger~\cite{str:07:RTA} that a proof in classical
propositional logic can be decomposed into a proof of (multiplicative)
linear logic, followed by a proof consisting only of contractions and
weakenings. In this paper we show that the same is possible for
first-order logic.

@@

A \bfit{substitution} is a function $\sigma\colon\VAR\to\TERM$ that is
the identity almost everywhere. We denote substitutions as
$\sigma=\sublist{\subst{x_1}{t_1},\ldots,\subst{x_n}{t_n}}$, where
$\sigma(x_i)=t_i$ for $i=1..n$ and $\sigma(x)=x$ for all
$x\notin\set{x_1,\ldots,x_n}$. 
We write $A\sigma$ for the formula
obtained from $A$ by applying $\sigma$, i.e., by simultaneously
replacing all occurrences of $x_i$ by $t_i$.
A \bfit{variable renaming} is a substitution $\rho$ with $\rho(x)\in\VAR$ for all variables $x$.

@@

\bfit{Sequents}, denoted by $\Gamma,\Delta,\ldots$, are finite
multisets of formulas, written as lists, separated by comma. The
\bfit{corresponding formula} of a (non-empty) sequent $\Gamma=A_1,A_2,\ldots,A_n$
is the disjunction of its formulas: $\form\Gamma=A_1\vlor
A_2\vlor\cdots\vlor A_n$. A sequent is \bfit{rectified} iff its
corresponding formula is.

@@

A cograph is \bfit{logical} if every vertex is labelled by either an
atom or variable, and it has at least one atom-labelled vertex. We
write $\single\alpha$ for an $\alpha$-labelled vertex. An
atom-labelled vertex is called a \bfit{literal} and a
variable-labelled vertex is called a \bfit{binder}. A binder labelled
with $x$ is called an \bfit{$x$-binder}. The \bfit{scope} of a binder
$b$ is the smallest proper strong module containing $b$.
An \bfit{$x$-literal} is a literal whose
atom contains the variable $x$. An $x$-binder \bfit{binds} every
$x$-literal in its scope.  In a logical cograph $\gG$, a binder $b$ is
\bfit{existential} (resp. \bfit{universal}) if, for every other vertex
$v$ in its scope, we have $bv \in \eG$ (resp. $bv \notin \eG$). An
$x$-binder is \bfit{legal} if its scope contains no other $x$-binder
and at least one literal.

\begin{definition}
A \bfit{first-order graph} or \bfit{fograph} is a logical cograph with legal binders. The
\bfit{binding graph} of a fograph $\gG$ is the directed graph $\bG=\tuple{\vG,
\set{(b, l) \mid b \text{ binds } l}}$.
\end{definition}

We define a mapping $\graphof\cdot$ from formulas to (labelled)
graphs, inductively as follows:
%% For a formula $A$, we can define its
%% associated fograph $\graphof{A}$ inductively by:\todo{duplicate two sentences above}
\begin{equation*}
  \begin{array}{r@{\;}l@{\hskip4em}r@{\;}l}
    \graphof{a} &= \single a \rlap{\hskip1em(for any atom $a$)}  \\[.9ex]
    \graphof{A \cor B} &= \graphof{A} + \graphof{B} &
    \graphof{\exists x . A} &= \single x \times \graphof{A} \\[.9ex]
    \graphof{A \cand B} &= \graphof{A} \times \graphof{B} &
    \graphof{\forall x . A} &= \single x + \graphof{A}    
  \end{array}
\end{equation*}
where $\single a$ (resp. $\single x$) is a single-vertex graph whose vertex is 
labelled by $a$ (resp. $x$).

@@

\begin{proof}
  That $\graphof A$ is a logical cograph follows immediately from the definition and Theorem~\ref{thm:cograph}.
  The fact that every binder of $\graphof A$ is legal can be proved by
  structural induction on $A$.
  %% :
  %% \begin{itemize}
  %%   \item Trivial if $A$ is an atom.
  %%   \item The induction step uses the following fact:
  %%         if $M$ is a module of $\gG$ and $M' \subseteq M$, then $M'$ is a
  %%         module of $\gG$ if and only if it is a module of $\gG[X]$.
  %% \end{itemize}
  %% \todo{}\lutz{maybe mention also fn 2-5 in Dominic's arxiv-paper}
\end{proof}
%%\juihsuan{I'm still looking for a simple proof}

\begin{remark}
Note that $\graphof A$ is not necessarily a fograph if $A$ is not
rectified. If $A = (\forall x. p(x)) \vlor (\forall x. q(x))$, then
$\graphof A = \single x \ \single p(x) \ \single x \ \single q(x)$,
the scope of each $x$-binder contains all the vertices, in particular,
the two $x$-binders. On the other hand, there are non-rectified
formulas which are translated to fographs by $\graphof\cdot$. For
example, in the graph of $(\exists x. p(x)) \vlor (\exists x. q(x))$,
both $x$-binders are legal, as they are not in each other's
scope:
$\edgepairpic{x}{px}\mkern9mu\edgepairpic{x}{qx}$.
\end{remark}

%
%% For this reason, we call a formula
%% \bfit{clean} if it does not contain subformulas of the form $(\forall
%% x.A)\vlor(\forall x.B)$ or $(\exists x.A)\vlan(\exists x.B)$,
%% \todo{need the four cases as in clear definition in fopws? Consider
%% $(\forall x.A)\vlor(B\vlor (\forall x.C))$. But then Lemma 6 doesn't hold.
%% (fopws proves only that there's a surjection from clear formulas onto fographs,
%% whose kernel is the six formula equalities)} and no
%% $x$-quantified formula occurs as subformula of another $x$-quantified
%% formula. Then we have:

%% \begin{lemma}
%%   If $A$ is clean iff\/ $\graphof A$ is a fograph.
%% \end{lemma}

%% \begin{proof}
%%   Induction on $A$, using  Theorem~\ref{thm:cograph}.
%% \end{proof}

%% Even though for every formula $A$ we can obtain an
%% equivalent clean formula by simply renaming some bound variables, this
%% is not unique up to $\alpha$-conversion, as is the case for
%% rectified formulas.

We define a congruence relation $\fequ$ on formulas by the following equations:
\begin{equation}
  \label{eq:fequ}
  \begin{array}{c@{\hskip1.5em}c}
    A \vlan B \fequ B \vlan A
    &
    (A \vlan B) \vlan C \fequ A \vlan (B \vlan C) 
    \\
    A \vlor B \fequ B \vlor A
    &
    (A \vlor B) \vlor C \fequ A \vlor (B \vlor C)
    \\
    \forall x . \forall y . A \fequ \forall y . \forall x . A
    &
    \forall x . (A \vlor B) \fequ (\forall x . A) \vlor B
    \\
    \exists x . \exists y . A \fequ \exists y . \exists x . A
    &
    \exists x . (A \vlan B) \fequ (\exists x . A) \vlan B
  \end{array}
\end{equation}
where $x \notin fv(B)$ in the last two equations.
Two formulas $A$ and $B$ are \bfit{equivalent} if $A \fequ B$. The following
theorem shows that the set of fographs can been seen as the quotient
$\FORM/\mathord\fequ$.
\begin{thm}
  Let $A,B$ be rectified formulas. Then
  $$
  A\fequ B \iff \graphof A =\graphof B
  $$
\end{thm}

@@

\begin{example}
  Both $\veedrinkerformula$ and $\variantveedrinkerformula$ (which are
  equivalent modulo $\fequ$) have the same (rectified) fograph
  $\drinkergraph$, shown below-left.
  %
  \begin{equation*}
      \drinkergraphpic
      %
      \hspace{8ex}
      %
      \drinkercotreepic
      %
      \hspace{8ex}
      %
      \drinkerbindinggraphpic
  \end{equation*}
  In the middle above we show the \emph{cotree} of the underlying
  cograph (illustrating the idea behind Theorem~\ref{thm:cograph}) and
  on the right above is its binding graph $\bD$.
\end{example}

@@

Let $\gC$ be a linked fograph. The set of links can be seen as a unification problem
by identifying dual predicate symbols. A \bfit{dualizer} of $\gC$ is a substitution $\delta$
unifying all the links of~$\gC$. Since a first-order unification problem is either
unsolvable or has a most general unifier, we can define the notion of \bfit{most
general dualizer}. A \bfit{dependency} is a pair $\set{\single x, \single y}$ of an
existential binder $\single x$ and a universal binder $\single y$ such that the most
general dualizer assigns to $x$ a term containing $y$. A \bfit{leap} is either a
link or a dependency. The \bfit{leap graph} $\lgC$ of $\gC$ is the undirected
graph $\tuple{\vC, \lpC}$ where $\lpC$ is the set of leaps of $\gC$. A vertex set $W \subseteq \vC$ induces a \bfit{matching} in
$\gC$ if for all $w \in W$, $N(w) \cap W$ is a singleton. We say that $W$ induces a
\bfit{bimatching} in $\gC$ if it induces a matching in $\gC$ and a matching in $\lgC$.

\begin{definition}
A \bfit{first-order net} or \bfit{fonet} is a linked fograph which has
dualizer but no induced bimatching.  
\end{definition}

\begin{figure}[!t]
  \begin{center}
    \vspace{-1ex}
    \begin{tikzpicture}[graph]
      \begin{scope}[xshift=-2.2cm]\twolinkfograph\end{scope}
        \begin{scope}[xshift=2.2cm]\twolinkleapgraph\end{scope}
    \end{tikzpicture}\vspace{-4ex}
  \end{center}%
  \caption{\label{fig:leap}A fonet (left) with
    dualizer $\protect\twolinkassignment$
    and its
    leap graph (right).}
\end{figure}

Figure~\ref{fig:leap} shows a fonet with a unique dualizer, and its leap graph.

@@

A graph homomorphism $\phi\colon \tuple{\vG, \eG} \rightarrow
\tuple{\vH, \eH}$ is a \bfit{fibration} if for all $v \in \vG$ and
$w\phi(v) \in \eH$, there exists a unique $\tilde{w} \in \vG$ such
that $\tilde{w}v \in \eG$ and $\phi(\tilde{w}) = w$ (indicated on the left below), and is a
\bfit{skew fibration} if for all $v \in \vG$ and $w\phi(v) \in \eH$
there exists $\tilde{w} \in \vG$ such that $\tilde{w}v \in \eG$ and
$\phi(\tilde{w})w \notin \eH$ (indicated in the middle below).  A directed graph homomorphism is a
\bfit{fibration} if for all $v \in \vG$ and $\pair{w,\phi(v)} \in
\eH$, there exists a unique $\tilde{w} \in \vG$ such that $\pair{\tilde{w},v}
\in \eG$ and $\phi(\tilde{w}) = w$ (indicated o the right below).
%
\liftingdiagrams

A \bfit{fograph homomorphism} $\phi=\tuple{\phi,\rsubstof\phi}$ is a
pair where $\phi\colon\gG\to\gH$ is a graph homomorphism between the
underlying graphs, and $\rsubstof\phi$, also called the \bfit{substitution induced by $\phi$} is a variable renaming such that
for all $v\in\vG$ we have
$\labelof{\phi(v)}=\rsubstof\phi(\labelof v)$. Note that this entails that
$\phi$ maps binders to binders and literals to literals.
We say that a
fograph homomorphism $\phi\colon\gG \rightarrow \gH$ is
%%\bfit{label-preserving} if for all $v \in \vG, label(v) = label(f(v))$, and is
\bfit{existential-preserving} if for all existential binders $b$ in
$\gG$, the vertex $\phi(b)$ is an existential binder in $\gH$.

\begin{definition}
  Let $\gG$ and $\gH$ be fographs. A \bfit{skew bifibration}
  $\phi\colon\gG\to\gH$ is an existential-preserving fograph
  homomorphism that is a skew fibration on $\tuple{\vG, \eG} \to
  \tuple{\vH, \eH}$ and a fibration on the binding graphs $\bG\to\bH$.
  %etween fographs is a label-preserving and
%existential-preserving fograph homomorphism whose restriction to the
%binding graphs is a fibration.
\end{definition}

\begin{example}
  Below on the left is a skew bifibration, with its binding fibration
  in the middle. We often omit the labeling in the upper graph, as
  done on the right below.
  $$
  \scalebox{.9}{
  \begin{tikzpicture}[graph]\drinkerfiblabelledpair{x}{x}
  \end{tikzpicture}
  \quad
  \begin{tikzpicture}[graph]\drinkerbindingfiblabelled{x}{x}
  \end{tikzpicture}
  \quad
  \begin{tikzpicture}[graph]\drinkerfib
  \end{tikzpicture}}
  $$
\end{example}

%%\subsection{First-order combinatorial proofs}

\begin{definition}
A \bfit{first-order combinatorial proof} (\bfit{FOCP}) of a fograph $\gG$ is a skew
bifibration $\phi\colon \gC \rightarrow \gG$ where $\gC$ is a fonet. A \bfit{first-order
combinatorial proof} of a formula $A$ is a combinatorial proof of its graph
$\graphof{A}$.
\end{definition}

To give some examples, we show in Figure~\ref{fig:cps} for examples of
FOCPs (taken from~\cite{hughes:fopws}), each above the formula it proves.  The same FOCPs are shown in
Figure~\ref{fig:cps-condensed} in a more condensed notation.

\begin{thm}[\cite{hughes:fopws}]
  \label{thm:FOCP}
  FOCPs are sound and complete for first-order logic.
\end{thm}

\begin{remark}
  In or definition of FOCP, we are slightly laxer than the original
  definition of~\cite{hughes:fopws}, as we allow for a variable
  renaming $\substof\phi$ which was forced to be the identity
  in~\cite{hughes:fopws}.
  %The reason for this relaxation in our case is that we want to cover a richer class of proofs, also allowing 
\end{remark}

@@

As with the sequent system $\FOLK$, we also need for $\FOKS$ the
\emph{linear fragment}, that we call here $\FOMLS$, and that is shown
in Figure~\ref{fig:KS1} in the dashed box.

We write $\vlder{\sysS}{\Deri}{A}{B}$ to denote a derivation $\Deri$
from $B$ to $A$ using the rules from system $\sysS$. A formula $A$ is
\bfit{provable} in a system $\sysS$ if there is a derivation in
$\sysS$ from $\ttt$ to $A$.

In the course of this paper we are also going to make use of the
general (non-atomic) version of the contraction rule:
\begin{equation*}
  \vlinf{\cD}{}{\Scons{A}}{\Scons{A\vlor A}}  
\end{equation*}

@@

We are now ready to see the main results of this paper. We only state
them here and give the proofs in the later sections of the paper. The
first one is routine and expected, but needs to be proved nonetheless:

>>>>

\newcommand{\compl}[1]{#1^\complement}
%\newcommand{\sublist}[1]{\set{#1}}
%\newcommand{\subst}[2]{#1\mathord{\scriptstyle\mapsto} #2}
\newcommand{\sublist}[1]{[#1]}
\newcommand{\subst}[2]{#1/#2}
\newcommand{\ssubst}[2]{\sublist{\subst{#1}{#2}}}
\newcommand{\substof}[1]{\sigma_{\!#1}}
\newcommand{\rsubstof}[1]{\rho_{#1}}
\newcommand{\dsubstof}[1]{\delta_{#1}}
\newcommand{\linkingof}[1]{\sim_{#1}}
\newcommand{\linking}{\sim}
\newcommand{\mapof}[1]{\lfloor{#1}\rfloor}
\newcommand{\labelof}[1]{\ell(#1)}
%\newcommand{\form}[1]{\mathrm{fm}(#1)}
\newcommand{\form}[1]{\bigvee\mkern-1mu(\mkern-2mu #1\mkern-2mu)}
\newcommand{\formtree}[1]{\mathcal{F}(#1)}

@@

In this paper we advance the idea that combinatorial proofs can provide
a notion of proof identity for first-order logic. \emph{First-order
combinatorial proofs} were introduced by Hughes in~\cite{hughes:fopws}.
For example, a first-order combinatorial proof of Smullyan's 
``drinker paradox'' $\drinkerformula=\veedrinkerformula$ is shown on the
right of Fig.\,\ref{fig:translate},\todo{wrong implies symbol in drinkerformula?}
a homomorphism from a partially coloured graph to a labelled graph.
However, even though Hughes proves soundness and completeness, the 
proof is highly unsatisfactory: (1) the soundness argument is extremely
long, intricate and cumbersome, and (2) the completeness proof does not 
allow a syntactic proof to be read back from a combinatorial proof, i.e.,
completeness is not
\emph{sequentializable} \cite{girard:87} or
\emph{full}~\cite{abramsky:jagadeesan:94}.\dominic{ok to say sequentializable too?}
%
A fundamental problem is that not all combinatorial
proofs can be obtained as translations of sequent calculus proofs.

We solve these issues by moving to a deep inference
system. More precisely, we introduce a new proof system, called
$\FOKS$, for first-order logic, that (a) reflects every combinatorial
proof, i.e., there is a surjection from $\FOKS$ proofs to
combinatorial proofs, (b) yields far simpler proofs of
soundness and completeness for combinatorial proofs, and (c) admits
new decomposition theorems establishing a precise correspondence
between certain syntactic inference rules and certain combinatorial
notions.
%
The right side of Fig.\,\ref{fig:translate}
illustrates the surjection in (a), and since the syntactic proofs 
of the two systems both translate the same combinatorial proof,
they can be considered \emph{the same}.

In general, a \emph{decomposition theorem} provides normal forms of
proofs, separating subsets of inference rules of a proof system. A
prominent example of a decomposition theorem is Herbrand's
theorem~\cite{herbrand:phd}, which allows a separation between the
propositional part and the quantifier part in a first-order
proof~\cite{gentzen:35:II,brunnler:06:herbrand}. Through the advent of
deep inference, new kinds of proof decompositions became possible,
most notably the separation between the linear part of a proof and the
resource management of a proof. It has been shown by
Stra{\ss}burger~\cite{str:07:RTA} that a proof in classical
propositional logic can be decomposed into a proof of multiplicative
linear logic, followed by a proof consisting only of contractions and
weakenings. In this paper we show that the same is possible for
first-order logic.

@@

A \bfit{substitution} is a function $\sigma\colon\VAR\to\TERM$ that is
the identity almost everywhere. We denote substitutions as
$\sigma=\sublist{\subst{x_1}{t_1},\ldots,\subst{x_n}{t_n}}$, where
$\sigma(x_i)=t_i$ for $i=1..n$ and $\sigma(x)=x$ for all
$x\notin\set{x_1,\ldots,x_n}$. 
Write $A\sigma$ for the formula
obtained from $A$ by applying $\sigma$, i.e., by simultaneously
replacing all occurrences of $x_i$ by $t_i$.
A \bfit{variable renaming} is a substitution $\rho$ with $\rho(x)\in\VAR$ for all variables $x$.

@@

\bfit{Sequents}, denoted by $\Gamma,\Delta,\ldots$, are finite
multisets of formulas, written as lists, separated by comma. The
\bfit{corresponding formula} of a (non-empty) sequent $\Gamma=A_1,A_2,\ldots,A_n$
is the disjunction of its formulas: $\form\Gamma=A_1\vlor
A_2\vlor\cdots\vlor A_n$. A sequent is \bfit{rectified} iff its
corresponding formula is.\dominic{I changed fm$(\Gamma)$ to $\form\Gamma$,
easier for referee to understand/intuit/remember later in the paper}

@@

A cograph is \bfit{logical} if every vertex is labelled by either an
atom or variable, and it has at least one atom-labelled vertex. We
write $\single\alpha$ for an $\alpha$-labelled vertex. An
atom-labelled vertex is called a \bfit{literal} and a
variable-labelled vertex is called a \bfit{binder}. A binder labelled
with $x$ is called an \bfit{$x$-binder}. The \bfit{scope} of a binder
$b$ is the smallest proper strong module containing $b$.
An \bfit{$x$-literal} is a literal whose
atom contains the variable $x$. An $x$-binder \bfit{binds} every
$x$-literal in its scope.  In a logical cograph $\gG$, a binder $b$ is
\bfit{existential} (resp. \bfit{universal}) if, for every other vertex
$v$ in its scope, we have $bv \in \eG$ (resp. $bv \notin \eG$). An
$x$-binder is \bfit{legal} if its scope contains no other $x$-binder
and at least one literal.
%
\begin{definition}
A \bfit{first-order graph} or \bfit{fograph} is a logical cograph with legal binders. The
\bfit{binding graph} of a fograph $\gG$ is the directed graph $\bG=\tuple{\vG,
\set{(b, l) \mid b \text{ binds } l}}$.
\end{definition}
%
\noindent We define a mapping $\graphof\cdot$ from formulas to (labelled)
graphs, inductively as follows:
%% For a formula $A$, we can define its
%% associated fograph $\graphof{A}$ inductively by:\todo{duplicate two sentences above}
\begin{equation*}
  \begin{array}{r@{\;}l@{\hskip4em}r@{\;}l}
    \graphof{a} &= \single a \rlap{\hskip1em(for any atom $a$)}  \\[.9ex]
    \graphof{A \cor B} &= \graphof{A} + \graphof{B} &
    \graphof{\exists x . A} &= \single x \times \graphof{A} \\[.9ex]
    \graphof{A \cand B} &= \graphof{A} \times \graphof{B} &
    \graphof{\forall x . A} &= \single x + \graphof{A}    
  \end{array}
\end{equation*}
where $\single a$ (resp. $\single x$) is a single-vertex graph whose vertex is 
labelled by $a$ (resp. $x$).

@@

\begin{proof}
  That $\graphof A$ is a logical cograph follows immediately from the definition and Theorem~\ref{thm:cograph}.
  The fact that every binder of $\graphof A$ is legal can be proved by
  structural induction on $A$.
  %% :
  %% \begin{itemize}
  %%   \item Trivial if $A$ is an atom.
  %%   \item The induction step uses the following fact:
  %%         if $M$ is a module of $\gG$ and $M' \subseteq M$, then $M'$ is a
  %%         module of $\gG$ if and only if it is a module of $\gG[X]$.
  %% \end{itemize}
  %% \todo{}\lutz{maybe mention also fn 2-5 in Dominic's arxiv-paper}
\end{proof}
%%\juihsuan{I'm still looking for a simple proof}
%
\begin{remark}
Note that $\graphof A$ is not necessarily a fograph if $A$ is not
rectified. If $A = (\forall x. p(x)) \vlor (\forall x. q(x))$, then
$\graphof A = \single x \ \single p(x) \ \single x \ \single q(x)$,
the scope of each $x$-binder contains all the vertices, in particular,
the other $x$-binder. On the other hand, there are non-rectified
formulas which are translated to fographs by $\graphof\cdot$. For
example, in the graph of $(\exists x. p(x)) \vlor (\exists x. q(x))$,
both $x$-binders are legal, as they are not in each other's
scope:
$\edgepairpic{x}{px}\mkern9mu\edgepairpic{x}{qx}$.
\end{remark}
%
%% For this reason, we call a formula
%% \bfit{clean} if it does not contain subformulas of the form $(\forall
%% x.A)\vlor(\forall x.B)$ or $(\exists x.A)\vlan(\exists x.B)$,
%% \todo{need the four cases as in clear definition in fopws? Consider
%% $(\forall x.A)\vlor(B\vlor (\forall x.C))$. But then Lemma 6 doesn't hold.
%% (fopws proves only that there's a surjection from clear formulas onto fographs,
%% whose kernel is the six formula equalities)} and no
%% $x$-quantified formula occurs as subformula of another $x$-quantified
%% formula. Then we have:
%
%% \begin{lemma}
%%   If $A$ is clean iff\/ $\graphof A$ is a fograph.
%% \end{lemma}
%
%% \begin{proof}
%%   Induction on $A$, using  Theorem~\ref{thm:cograph}.
%% \end{proof}
%
%% Even though for every formula $A$ we can obtain an
%% equivalent clean formula by simply renaming some bound variables, this
%% is not unique up to $\alpha$-conversion, as is the case for
%% rectified formulas.
%
\noindent We define a congruence relation $\fequ$ on formulas by the following equations:
\begin{equation}
  \label{eq:fequ}
  \begin{array}{c@{\hskip1.5em}c}
    A \vlan B \fequ B \vlan A
    &
    (A \vlan B) \vlan C \fequ A \vlan (B \vlan C) 
    \\
    A \vlor B \fequ B \vlor A
    &
    (A \vlor B) \vlor C \fequ A \vlor (B \vlor C)
    \\
    \forall x . \forall y . A \fequ \forall y . \forall x . A
    &
    \forall x . (A \vlor B) \fequ (\forall x . A) \vlor B
    \\
    \exists x . \exists y . A \fequ \exists y . \exists x . A
    &
    \exists x . (A \vlan B) \fequ (\exists x . A) \vlan B
  \end{array}
\end{equation}
where $x \notin fv(B)$ in the last two equations.
\todo{Define $fv$; used later too, e.g.\ proof of Lem.\,\ref{lem:me}}
Two formulas $A$ and $B$ are \bfit{equivalent} if $A \fequ B$. The following
theorem shows that the set of fographs can been seen as the quotient
$\FORM/\mathord\fequ$.
\begin{thm}
  Let $A,B$ be rectified formulas. Then
  $$
  A\fequ B \iff \graphof A =\graphof B
  $$
\end{thm}

@@

\begin{example}
  Both $\veedrinkerformula$ and $\variantveedrinkerformula$, which are
  equivalent modulo $\fequ$, have the same (rectified) fograph
  $\drinkergraph$, shown below-left.
  %
  \begin{equation*}
      \drinkergraphpic
      %
      \hspace{8ex}
      %
      \drinkercotreepic
      %
      \hspace{8ex}
      %
      \drinkerbindinggraphpic
  \end{equation*}
  Above-left we show the \emph{cotree} of the underlying
  cograph (illustrating the idea behind Theorem~\ref{thm:cograph}) and
  above-right is its binding graph $\bD$.
\end{example}

@@

Let $\gC$ be a linked fograph. The set of links can be seen as a unification problem
by identifying dual predicate symbols. A \bfit{dualizer} of $\gC$ is a substitution $\delta$
unifying all the links of~$\gC$. Since a first-order unification problem is either
unsolvable or has a most general unifier, we can define the notion of \bfit{most
general dualizer}. A \bfit{dependency} is a pair $\set{\single x, \single y}$ of an
existential binder $\single x$ and a universal binder $\single y$ such that the most
general dualizer assigns to $x$ a term containing $y$. A \bfit{leap} is either a
link or a dependency. The \bfit{leap graph} $\lgC$ of $\gC$ is the undirected
graph $\tuple{\vC, \lpC}$ where $\lpC$ is the set of leaps of $\gC$.
A vertex set $W \subseteq \vC$ induces a \bfit{matching} in
$\gC$ if 
it is non-empty\dominic{added non-empty (necessary, right?)}
for all $w \in W$, $N(w) \cap W$ is a singleton. We say that $W$ induces a
\bfit{bimatching} in $\gC$ if it induces a matching in $\gC$ and a matching in $\lgC$.

\begin{definition}
A \bfit{first-order net} or \bfit{fonet} is a linked fograph which
has a dualizer but no induced bimatching.  
\end{definition}
%
\begin{figure}[!t]
  \begin{center}
    \vspace{-1ex}
    \begin{tikzpicture}[graph]
      \begin{scope}[xshift=-2.2cm]\twolinkfograph\end{scope}
        \begin{scope}[xshift=2.2cm]\twolinkleapgraph\end{scope}
    \end{tikzpicture}\vspace{-4ex}
  \end{center}%
  \caption{\label{fig:leap}A fonet (left) with
    dualizer $\protect\twolinkassignment$
    and its
    leap graph (right).}
\end{figure}%
%
Figure~\ref{fig:leap} shows a fonet with a unique dualizer, and its leap graph.

@@

A graph homomorphism $\phi\colon \tuple{\vG, \eG} \rightarrow
\tuple{\vH, \eH}$ is a \bfit{fibration} if for all $v \in \vG$ and
$w\phi(v) \in \eH$, there exists a unique $\tilde{w} \in \vG$ such
that $\tilde{w}v \in \eG$ and $\phi(\tilde{w}) = w$ (indicated below-left),
and is a
\bfit{skew fibration} if for all $v \in \vG$ and $w\phi(v) \in \eH$
there exists $\tilde{w} \in \vG$ such that $\tilde{w}v \in \eG$ and
$\phi(\tilde{w})w \notin \eH$ (indicated below-centre).
A directed graph homomorphism is a
\bfit{fibration} if for all $v \in \vG$ and $\pair{w,\phi(v)} \in
\eH$, there exists a unique $\tilde{w} \in \vG$ such that $\pair{\tilde{w},v}
\in \eG$ and $\phi(\tilde{w}) = w$ (indicated below-right).
%
\liftingdiagrams
%
A \bfit{fograph homomorphism} $\phi=\tuple{\phi,\rsubstof\phi}$ is a
pair where $\phi\colon\gG\to\gH$ is a graph homomorphism between the
underlying graphs, and $\rsubstof\phi$, also called the
\bfit{substitution induced by $\phi$} is a variable renaming 
such that
for all $v\in\vG$ we have
$\labelof{\phi(v)}=\rsubstof\phi(\labelof v)$,
and $\rsubstof\phi$ is the identity on variables not in $\gG$.
%
Note that $\phi$ necessarily maps binders to binders
and literals to literals.
%
Since $\rsubstof\phi$ is fully determined by $\phi$ alone, we often
leave $\rsubstof\phi$ implicit.
A fograph homomorphism $\phi\colon\gG \rightarrow \gH$
%%\bfit{label-preserving} if for all $v \in \vG, label(v) = label(f(v))$, and is
% is \bfit{existential-preserving} 
\bfit{preserves existentials}
if for all existential binders $b$ in
$\gG$, the binder $\phi(b)$ is a existential in $\gH$.\dominic{tweaked this paragraph}
%
\begin{definition}
  Let $\gG$ and $\gH$ be fographs. A \bfit{skew bifibration}
  $\phi\colon\gG\to\gH$ is an existential-preserving fograph
  homomorphism that is a skew fibration on $\tuple{\vG, \eG} \to
  \tuple{\vH, \eH}$ and a fibration on the binding graphs $\bG\to\bH$.
  %etween fographs is a label-preserving and
%existential-preserving fograph homomorphism whose restriction to the
%binding graphs is a fibration.
\end{definition}
%
\begin{example}
  Below-left is a skew bifibration, whose binding fibration
  is below-centre. When the labels on the source fograph can be inferred
  (modulo renaming),
  we often omit the labeling in the upper graph, as below-right, which refer to as 
  \bfit{condensed form}.\dominic{I said ``can be inferred'' to account for 
  e.g.\
  $\graphof{\forall x_1\forall x_2 (px_i\vee px_j\vee px_k\vee px_l)} \to \forall x px$ whose
  unlabelled skeleton is the skeleton of many distinct formula variants (``who binds who?'').}
  \begin{center}\begin{math}
  \scalebox{.9}{
  \begin{tikzpicture}[graph]\drinkerfiblabelledpair{x}{x}
  \end{tikzpicture}
  \quad
  \begin{tikzpicture}[graph]\drinkerbindingfiblabelled{x}{x}
  \end{tikzpicture}
  \quad
  \begin{tikzpicture}[graph]\drinkerfib
  \end{tikzpicture}}
\end{math}\end{center}\end{example}
%
%%\subsection{First-order combinatorial proofs}
%
\begin{definition}
A \bfit{first-order combinatorial proof} (\bfit{FOCP}) of a fograph $\gG$ is a skew
bifibration $\phi\colon \gC \rightarrow \gG$ where $\gC$ is a fonet. A \bfit{first-order
combinatorial proof} of a formula $A$ is a combinatorial proof of its graph
$\graphof{A}$.
\end{definition}
%
Figure~\ref{fig:cps} shows examples of
FOCPs (taken from~\cite{hughes:fopws}), each above the formula it proves.
The same FOCPs are shown in
Figure~\ref{fig:cps-condensed} in condensed form.
%
\begin{thm}[\cite{hughes:fopws}]
  \label{thm:FOCP}
  FOCPs are sound and complete for first-order logic.
\end{thm}
%
\begin{remark}
  Our definition of FOCP is slightly more lax than the original
  definition of~\cite{hughes:fopws}, as we allow for a variable
  renaming $\substof\phi$ which was forced to be the identity
  in~\cite{hughes:fopws}.
  %The reason for this relaxation in our case is that we want to cover a richer class of proofs, also allowing 
\end{remark}

@@

As with the sequent system $\FOLK$, we also need for $\FOKS$ the
\emph{linear fragment}, $\FOMLS$, and that is shown
in Figure~\ref{fig:KS1} in the dashed box.

Write $\vlder{\sysS}{\Deri}{A}{B}$ to denote a derivation $\Deri$
from $B$ to $A$ using the rules from system $\sysS$. A formula $A$ is
\bfit{provable} in a system $\sysS$ if there is a derivation in
$\sysS$ from $\ttt$ to $A$.

In the course of this paper we will employ the
general (non-atomic) version of the contraction rule:
\begin{equation*}
  \vlinf{\cD}{}{\Scons{A}}{\Scons{A\vlor A}}  
\end{equation*}

@@

We state the main results of this paper here,
and prove them in later sections. The
first is routine and expected, but needs to be proved nonetheless:
